Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-28 paper_jasa.tex
pvalues in real world updated

\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?} \blue{$p$-values are computed fitting the full model to the contaminated data anyway. So many small p-values are coming from the fact that contamination makes the slope negatively significant. For example, for some slope the contamination reverses the slope (positive to negative) and negative slope got significant. It is like the effect of influential outlier.}

In the real setting, where visual inference is to be useful, there will be no conventional tests or traditional $p$-values. Yet, assessing the strength of the structure is a critical component of inference. For experiments 1 and 2, to assess the signal strength we examine the proportion of correct picks, with multiple users and the $p$-value of the actual data plot. For experiment 3 the \green{How is p-value calculated here?} Figure \ref{fig:pval_pcorrect} \blue{(should I bring this figure back?)}  displays scatterplots of these two variables for each experiment. As the $p$-value increases the proportion of correct responses falls, evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is almost impossible to correctly identify the observed plot in the lineup. We observed this for both experiment 1 and 2. For experiment 3, the scenario is different, because there are two structures, the association between variables, and the cluster of contaminated points.% as in this case people might have observed something other than the effect $E$.

\green{Should we use proportion of correct, rather than percentage? Wouldn't that work into the equations better?}

Yet, it is possible to obtain an estimate the strength of the signal in the actual data plot, when there are multiple evaluations, using Lemma \ref{lemma}. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. 
Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values. 
\hh{Note that the estimate for $p_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the  choice of a large lineup size, $m$, will make the value look `significant' in a traditional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.}
\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.48, 0.85, and 0.11 based on $n_1 = 39, n_2 = 41$, and $n_3= 18$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to classical test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

>>>>

%\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?} \blue{$p$-values are computed fitting the full model to the contaminated data anyway. So many small p-values are coming from the fact that contamination makes the slope negatively significant. For example, for some slope the contamination reverses the slope (positive to negative) and negative slope got significant. It is like the effect of influential outlier.}

In the real setting, where visual inference is to be useful, there will be no conventional tests or traditional $p$-values. Assessing the strength of perceived structure is a critical component of visual inference. In experiments 1 and 2, there is a $p$-value associated with the actual data plot in each lineup. As the $p$-value increases the proportion of correct responses falls (Figure \ref{fig:pval_pcorrect}), which is evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is very uncommon for subjects to correctly identify the actual data plot in the lineup.

\green{Add figure showing proportion correct vs p-value of actual data plot for experiments 1 and 2.}

\green{You need to explain visual $p$-value - how is this calculated. How is this different from the signal strength. And please say why you introduce a term signal strength.}

With multiple observers it is possible to estimate the strength of the signal, using the proportion of correct picks. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?}

\green{Be consistent with capitalization of plot labels - check this everywhere - you have lower case signal and upper case Strength}

%, to assess the signal strength we examine the proportion of correct picks, with multiple users and the $p$-value of the actual data plot. For experiment 3 the \green{How is p-value calculated here?} Figure \ref{fig:pval_pcorrect} \blue{(should I bring this figure back?)}  displays scatterplots of these two variables for each experiment. As the $p$-value increases the proportion of correct responses falls, evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is almost impossible to correctly identify the observed plot in the lineup. We observed this for both experiment 1 and 2. For experiment 3, the scenario is different, because there are two structures, the association between variables, and the cluster of contaminated points.% as in this case people might have observed something other than the effect $E$.

%Yet, it is possible to obtain an estimate the strength of the signal in the actual data plot, when there are multiple evaluations, using Lemma \ref{lemma}.

Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a traditional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.48, 0.85, and 0.11 based on $n_1 = 39, n_2 = 41$, and $n_3= 18$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to classical test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.
